Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex
Move to {p|l.r} syntax in 4.0 to be consistent with rest of document.

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $p$ and $l.r$. ($p$ and $l.r$
contain all first and second stage computations in $e$ respectively.)
Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
$v$.  Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms $c$
and $l.r$. Evaluating $c$ in the first pass produces the pair $(i,b)$,
where the first element $i$ matches the first stage results contained
in $v$.  For example, in the simple case where $\rtab \xi v$ contains
entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is used as input to the resumer in second pass evaluation as $\reduce
{[b/l]r} {v'}$, where $v'$ is the same as result produced by second
stage computations in $e$.  For example, in the case where $\rtab \xi
v$ contains entirely second stage computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).

@@

\subsection {Necessity of \bbonep}
\label{sec:needGround}

We have not yet given a justification for why first stage code must be
partitioned between the worlds \bbonem\ and \bbonep.
Certainly one could imagine a simpler system without all the \texttt{grnd} annotations.  
Why wouldn't this work?

@@

We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
Yet even if a language designer desired a different implicit/explicit trade-off,
then the three-world \lang\ would still be useful as a typed intermediate representation.

>>>>

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms {$p$|$l.r$}. ($p$ and $l.r$
contain all first and second stage computations in $e$ respectively.)
Executing $p$ in a first pass produces $b$ (via $\reduce p b$). Then,
in the second pass, $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
$v$.  Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
{$c$|$l.r$}. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into stage one and stage two results. We then describe
the algorithm for splitting both \bbonem\ terms ($\splitonesym$) and
\bbtwo\ terms ($\splittwosym$).

@@

\subsection {Role of World \bbonep}
\label{sec:needGround}

The splitting algorithm described in the previous subsections is
defined based on the local structure of \lang\ terms.  One artifact of
this design is that splitting generates

For example, the rule for splitting \bbonem\ case terms

% splitting rules are local term driven

% generate a lot of cruft

% for example, case pushes the state of the case through to 2 even if branches have no stage 1 content

% recursive case would be recursive calls

% Rather than try to optimize globally So we use the type system to prove statically there no stage 1 stuff

% skip splitting.

% example is partition in 6.3

@@

%We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
%Yet even if a language designer desired a different implicit/explicit trade-off,
%then the three-world \lang\ would still be useful as a typed intermediate representation.
